---
title: "Proposal title"
subtitle: "Proposal"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
categories: 
  - Data visualization
execute:
  warning: false
---

```{r load_pkgs, message=FALSE, output=FALSE}
#| code-fold: true
#| code-summary: "Installed Packages"

# GETTING THE LIBRARIES
if (!require(pacman))
  install.packages(pacman)

pacman::p_load(tidyverse,
               formattable,
               tidytuesdayR,
               janitor,
               dlookr,
               here)
```

## Dataset

```{r load_dataset, message=FALSE, output=FALSE}
# Reading the data using read_csv
flights_ntsb <- read_csv(here("data", "flight_crash_data_NTSB.csv"))

flights_ntsb <- flights_ntsb |>
  clean_names()
```

## Background

We had to submit a formal request to the National Transportation Safety Board (NSTB) to obtain the raw dataset. This dataset is for the United States of America only. The data dimensions is a tibble of 89,134 rows and 38 columns, culminating 3,387,093 variables. The data goes back to 1-1-1980 until 12-31-2022.

```{r diagnose_flights, message=FALSE}
#| code-fold: true
#| code-summary: "Flights Report Data Diagnosis Code"

# Getting basic information about the dataset

flights_ntsb |>
  diagnose() |>
  formattable()
```

```{r describe_flights, message=FALSE}
#| code-fold: true
#| code-summary: "Flights Report Data Describe Code"

# Getting basic statistical information about the dataset

flights_ntsb |>
  describe() |>
  formattable()
```

## Value

After talking about places we've traveled, we chose the data as some of us are more comfortable with flying than others. We decided by looking at aircraft crash data that it could be a chance to get more informed about aircraft crashes within the United States. We wanted do timeseries analysis as well as geospatial analysis and this data presented itself as a great opportunity to do both.

## Questions

The two questions to be answered are:

**Question 1**

How have the number of crashes and injuries evolved over time?

Time-based Analysis -Visualization: Line graphs with "Event date" on the x-axis, stacked time series count showing highest injury count, and counts of crashes or injuries on the y-axis. This can help in visualizing trends over time.

**Question 2**

Where do most aircraft crashes occur?

Geographical Analysis-Visualization: A geographical heatmap or scatter plot using latitude and longitude to pinpoint crash locations. Areas with a high concentration of crashes will be highlighted.

## Analysis plan

### **Approach for question 1**

### **Approach for question 2**

data cleaning and data wrangling - new columns and removing columns - removing NAs,

weather condition for q2 - rayshader as a possibility to display

shiny app - how will we use to display - year or decade

weekly plan of attack -

## Implementation

### Plan of Attack

### Repo Organization
