---
title: "Comprehensive analysis of air crash data"
subtitle: "Proposal"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
categories: 
  - Data visualization
execute:
  warning: false
---

```{r load_pkgs, message=FALSE, output=FALSE}
#| code-fold: true
#| code-summary: "Installed Packages"

# GETTING THE LIBRARIES
if (!require(pacman))
  install.packages(pacman)

pacman::p_load(tidyverse,
               formattable,
               tidytuesdayR,
               janitor,
               dlookr,
               here)
```

## Dataset

```{r load_dataset, message=FALSE, output=FALSE}
# Reading the data using read_csv
flights_ntsb <- read_csv(here("data", "flight_crash_data_NTSB.csv"))

flights_ntsb <- flights_ntsb |>
  clean_names()
```

## Background

We had to submit a query to the National Transportation Safety Board ([NSTB](https://www.ntsb.gov/Pages/AviationQueryV2.aspx)) to obtain the raw dataset. This dataset is for the United States of America only. The data dimensions is a tibble of 89,134 rows and 38 columns, culminating 3,387,093 variables. The data goes back to 1-1-1980 until 12-31-2022.

```{r diagnose_flights, message=FALSE}
#| code-fold: true
#| code-summary: "Flights Report Data Diagnosis Code"

# Getting basic information about the dataset

flights_ntsb |>
  diagnose() |>
  formattable()
```

```{r describe_flights, message=FALSE}
#| code-fold: true
#| code-summary: "Flights Report Data Describe Code"

# Getting basic statistical information about the dataset

flights_ntsb |>
  describe() |>
  formattable()
```

## Value

After talking about places we've traveled to, we chose the data as some of us are more comfortable with flying than others. We decided by looking at aircraft crash data that it could be a chance to get more informed about aircraft crashes within the United States. We wanted do timeseries analysis as well as geospatial analysis and this data presented itself as a great opportunity to do both.

## Questions

The two questions to be answered are:

### Question 1

What is the historical trend in aircraft crashes and injuries, including the distribution of injury severity (fatal, serious, minor)? Additionally, which events lead to the most injuries or fatalities?

### Question 2

Where are the most common locations for aircraft crashes, and are there any specific connections like months, weather conditions associated with crash frequencies?

## Analysis plan

### Approach for question 1

Time-based Analysis -Visualization: Line graphs with "Event date" on the x-axis, stacked time series count showing highest injury count, and counts of crashes or injuries on the y-axis. This can help in visualizing trends over time.

### Approach for question 2

Geographical Analysis-Visualization: A geographical heatmap or scatter plot using latitude and longitude to pinpoint crash locations. Areas with a high concentration of crashes will be highlighted.

data cleaning and data wrangling - new columns and removing columns - removing NAs,

weather condition for q2 - rayshader as a possibility to display

shiny app - how will we use to display - year or decade

weekly plan of attack -

## Implementation

### Plan of Attack

|               Week                | Weekly Tasks                                                                          | Persons in Charge |  Backup  |
|:------------------:|---------------------------------|:----------------:|:-------------:|
|       until November 8^th^        | Explore and finalize the dataset and the problem statements                           |     Everyone      | Everyone |
|                \-                 | Complete the proposal and assign some high-level tasks                                |     Everyone      | Everyone |
|     November 9^th^ to 15^th^      | Getting to know about Shiny application                                               |        TBD        |   TBD    |
|                \-                 | Data cleaning and Data pre-processing                                                 |        TBD        |   TBD    |
|                \-                 | Question specific exploration and data categorization \<change to question specific\> |        TBD        |   TBD    |
|     November 16^th^ to 22^nd^     | Generating plots for Q1                                                               |        TBD        |   TBD    |
|                \-                 | Generating plots for Q2                                                               |        TBD        |   TBD    |
|                \-                 | Exploring on how to integrate our specific visualizations and shiny                   |        TBD        |   TBD    |
|     November 23^rd^ to 29^th^     | Generating remaining parts of the plots for Q1 and Q2                                 |        TBD        |   TBD    |
|                \-                 | Improving the generated plots                                                         |        TBD        |   TBD    |
|                \-                 | Start integrating shiny and our visualizations                                        |        TBD        |   TBD    |
| November 30^th^ to December 6^th^ | Refining the code for code review with comments                                       |        TBD        |   TBD    |
|                \-                 | Having some base functions to use themes interchangably                               |        TBD        |   TBD    |
|                \-                 | Continue with the integration of shiny and our plots                                  |        TBD        |   TBD    |
|     December 7^th^ to 13^th^      | Complete the shiny application with multiple user functionality                       |        TBD        |   TBD    |
|                \-                 | Review the generated plots and shiny integration                                      |        TBD        |   TBD    |
|                \-                 | Write-up and presentation for the project                                             |     Everyone      | Everyone |

: {.hover}

### Repo Organization

-   **'data/':** Used for storing any necessary data files for the project, such as input files.

-   **'images/':**

-   **'presentation_files/':**

-   **'\_extra/':**

-   **'\_freeze/':**

-   **'\_site/':**

-   **'.github/':**
