---
title: "radial-bar-plot"
author: "Eshaan Mathakari"
editor: visual
---

We will be representing the count of crashes on the x-axis and the phases of flight (takeoff, landing etc,.) on the y-axis in a circular manner. We can transform the bar plot into a radial form, emphasizing the distribution of crashes during these critical flight stages. This visualization technique allows for a quick understanding of the relative frequency of crashes during takeoff and landing, enabling insights into the safety challenges faced during these flight phases.
highcharter, ggiraph - interactive plot

```{r installing-packages}
pacman::p_load(tidyverse, 
               ggplot2, 
               here, 
               highcharter, 
               ggiraph,
               dplyr)
```

```{r new-dataset}
edited_data <- read_csv("https://raw.githubusercontent.com/INFO526-DataViz/project-final-Infographic-Innovators/main/data/edited-flights-data.csv")
```

```{r radial-plot}
flight_phases <- c("Cruise", "Maneuvering", "Approach", "Takeoff", "Climb",
                   "Descent", "Landing", "Hover", "Standing", "Other",
                   "Taxi", "Uncontrolled Descent", "Emergency", "Holding", "Unknown")

# Subset the data to include only the specified flight phases
filtered_data <- edited_data[edited_data$Phase %in% flight_phases, ]

filtered_data$Phase <- factor(filtered_data$Phase, levels = flight_phases)

radar_plot <- ggplot(filtered_data, aes(x = as.factor(Phase), y = ..count..)) +
  geom_bar(stat = "count", fill = "skyblue") +
  coord_polar() +
  theme_minimal() +
  labs(title = "Radial Bar Plot of Flight Phases and Crash Counts",
       x = "Flight Phase",
       y = "Number of Crashes occuring")

print(radar_plot)
```
```{r}
flight_phases <- c("Cruise", "Maneuvering", "Approach", "Takeoff", "Landing")

filtered_data <- edited_data[edited_data$Phase %in% flight_phases, ]

# Create a factor variable for 'Phase' with a custom order
filtered_data$Phase <- factor(filtered_data$Phase, levels = flight_phases)

# Define colors for each flight phase
phase_colors <- c("#00FA9A", "#9ACD32", "#FA8072", "#FF0000", "#DC143C")

radial_bar_plot <- ggplot(filtered_data, aes(x = as.factor(Phase), fill = Phase)) +
  geom_bar(width = 0.9, stat = "count") +
  coord_polar(theta = "y") +
  xlab("") + ylab("Count of Crashes") +
  ggtitle("Count of Crashes by Flight Phase") +
  geom_text(stat = "count", aes(label = ..count..),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = phase_colors, name = "Flight Phase",
                    labels = c("Cruise", "Maneuvering", "Approach", "Takeoff", "Landing")) +
  theme_minimal() +
  theme(legend.position = "bottom",  # Adjust legend position as needed
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(size = 8),
        axis.ticks = element_blank())

# Display the plot
print(radial_bar_plot)


```

